Efficient implementation of penalized regression\\for genetic risk prediction

### 2019-01-23 paper/PRS.tex
nearly last updates

{\color{red}
In this paper, we present an efficient method to jointly estimate SNP effects using individual-level data, allowing for practical application of penalized logistic regression (PLR) on modern datasets including hundreds of thousands of individuals. Moreover, our implementation of PLR directly includes automatic choices for hyper-parameters.
}

We compare the performance of PLR, C+T and a derivation of random forests using both real and simulated data. 
PLR consistently achieves higher predictive performance than the two other methods while being as fast as C+T. 
We find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; for instance, AUC values increase from 83\% with the best prediction of C+T to 92.5\% with PLR. We confirm these results in a data analysis of a case-control study for celiac disease where PLR and the standard C+T method achieve AUC of 89\% and of 82.5\%.

In conclusion, our study demonstrates that penalized logistic regression can achieve more discriminative polygenic risk scores, while being applicable to large-scale individual-level data thanks to the implementation we provide in the R package bigstatsr.
}

{\color{red}
Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being as fast. This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some well power studies in return. 
}

@@

To perform simulations, we use real genotype data and simulate new phenotypes. 
In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the disease heritability. We also compare two different models for simulating phenotypes, one with additive effects only, and one that combines additive, dominant and interaction-type effects.
{\color{red}
Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being as fast. This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some high predictive performance in return. 
}

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWAS). In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$. The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another (this set of SNPs is denoted $S_\text{clumping}$). Then, thresholding (T) consists in removing SNPs with p-values larger than a user-defined threshold $p_T$. Finally, the PRS for individual $i$ is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients \[\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,\] where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$. We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``C+T-all''), one including only the SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``C+T-stringent''), and one that maximizes the AUC (``C+T-max'') for 102 p-value thresholds between $1$ and $10^{-100}$ (Table \ref{tab:thr}).
As we report the optimal threshold based on the test set, the AUC for ``C+T-max'' is an upper bound of the AUC for the C+T method.
{\color{red}
For the computation of this paper, the GWAS part is done on the training set while clumping is done on the test set (all individuals not included in the training set).
}

@@

We tested our implementation on 656K genotyped SNPs of the UK Biobank, keeping only caucasian individuals and removing individuals related to someone else in the data (second individual in each pair with a kinship greater than 0.08). Our L1-penalized linear regression runs in less than one day for 350K individuals, achieving a correlation of 65.5\% with true height for both sexes. The best C+T model achieves a correlation of 55\% for women and 56\% for men, where the GWAS part takes 1 hour only.
Our L1-penalized logistic regression on breast cancer runs in only 12 minutes for 150K women, achieving an AUC of 0.587. The best C+T model achieves the same AUC score, where the GWAS part takes 15 hours.

[TODO: RESULTS TABLE]

@@

The computation time of our PLR implementation mainly depends on the sample size and the number of active variables. Indeed, the algorithm is composed of two steps: first, for each variable, the algorithm computes one statistic that will be used to decide if the variable enters the model or not. Then, the algorithm iterates over a regularization path of decreasing values of $\lambda$, which progressively enables variables to enter the model (Figure  \ref{fig:CMSA}). So, in the second step, the number of variables is increasing progressively, and at some point the computation stops due to the early stopping criterion (when prediction is getting worse on the corresponding validation set, see figure \ref{fig:CMSA}).

@@

Regarding memory requirements, this is basically the same as for computation time. Indeed, active variables are accessed in memory thanks to memory-mapping when they are used \cite[]{prive2017efficient}. 
If there are too many variables accessed, the operating system (OS) has to make some space for new coming variables. 
So, there should never be any memory issue thanks to memory-mapping. Yet, at some point, if there are too many active variables, the OS would regularly swap memory between disk and RAM, which is very slow. 
If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM. 
We have quickly tested processing each chromosome separately and then combining the models later but this did not seem to perform as well as fitting all SNPs at once (data not shown). 
Another widely used approach is to apply the penalized regression on a subset of SNPs using univariate tests (GWAS) to prioritize SNPs. Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm here by achieving a correlation of 65.5\% for height using 350K individuals of the UK Biobank, whereas \cite{accuheight2018} achieved only 61\% for the same sample size when using only the 50K most significantly associated SNPs (see figure 3 of their paper).

@@

When there are many effects of small size and a small sample size, PLR performs worse than (the best result for) C+T. This may be because PLR is too conservative in including effects, whereas it has been shown in the literature that, in such underpowered scenarios, including (nearly) all SNPs was the best strategy \cite[]{purcell2009common}.
In scenario \textnumero3, by varying sample size, we showed that when increasing sample size, PLR would achieve higher predictive performance than C+T at some point and that the benefits of using PLR over C+T would increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}). Moreover, in scenario \textnumero2, we decrease the search space (p: number of SNPs) in order to increase the polygenicity of the models and to virtually increase the sample size \cite[]{dudbridge2013power}. In this scenario, even when there are small effects and a high polygenicity (3000 causal variants out of 18,941), PLR gets a large increase in predictive performance, now consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
}

>>>>

{\color{red}
In this paper, we present an efficient method to jointly estimate SNP effects using individual-level data, allowing for practical application of penalized logistic regression (PLR) on modern datasets including hundreds of thousands of individuals. Moreover, our implementation of PLR directly includes automatic choices for hyper-parameters. We also provide a similar implementation of penalized linear regression.
}

We compare the performance of PLR, C+T and a derivation of random forests using both real and simulated data. 
{\color{red}
Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being scalable to biobank data.
}
We find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; for instance, in simulations, AUC values increase from 83\% with the best prediction of C+T to 92.5\% with PLR. We confirm these results in a data analysis of a case-control study for celiac disease where PLR and the standard C+T method achieve AUC of 89\% and of 82.5\%.

{\color{red}
Using 350,000 individuals of the UK Biobank, we predict height with 10\% more correlation than with the best prediction of C+T. Although height is a highly polygenic trait with many small effects, our implementation of penalized regression proves to be scalable and, to the best of our knowledge, provides the best genetic predictor of height to date.
Moreover, using 150,000 individuals of the UK Biobank, we are able to predict breast cancer better than C+T, fitting PLR in a few minutes only.
This demonstrates the feasibility and relevance of using penalized regression for PRS computation when large individual-level datasets are available.
}

In conclusion, our study demonstrates that penalized regression can achieve much more discriminative polygenic risk scores, while being applicable to large-scale individual-level data thanks to the implementation we provide in R package bigstatsr.
}

@@

To perform simulations, we use real genotype data and simulate new phenotypes. 
In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the disease heritability. We also compare two different models for simulating phenotypes, one with additive effects only, and one that combines additive, dominant and interaction-type effects.
{\color{red}
Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being scalable to biobank data. %This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some high predictive performance in return. 
}

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWAS). In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$. The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another (this set of SNPs is denoted $S_\text{clumping}$). Then, thresholding (T) consists in removing SNPs with p-values larger than a user-defined threshold $p_T$. Finally, the PRS for individual $i$ is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients \[\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,\] where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$. We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``C+T-all''), one including only the SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``C+T-stringent''), and one that maximizes the AUC (``C+T-max'') for 102 p-value thresholds between $1$ and $10^{-100}$ (Table \ref{tab:thr}).
As we report the optimal threshold based on the test set, the AUC for ``C+T-max'' is an upper bound of the AUC for the C+T method.
{\color{red}
For the computation of this paper, the GWAS part uses the training set while clumping uses the test set (all individuals not included in the training set).
}

@@

We tested our implementation on 656K genotyped SNPs of the UK Biobank, keeping only caucasian individuals and removing individuals related to someone else in the data (excluding the second individual in each pair with a kinship greater than 0.08). Our L1-penalized linear regression runs in less than one day for 350K individuals, achieving a correlation of 65.5\% with true height for both sexes. The best C+T model achieves a correlation of 55\% for women and 56\% for men, where the GWAS part takes 1 hour only.
Our L1-penalized logistic regression on breast cancer runs in only 12 minutes for 150K women, achieving an AUC of 0.587. The best C+T model achieves the same AUC score, where the GWAS part takes 15 hours.

[TODO: COMPLETER + RESULTS TABLE]

@@

The computation time of our PLR implementation mainly depends on the sample size and the number of active variables. Indeed, the algorithm is composed of two steps: first, for each variable, the algorithm computes one statistic that will be used to decide if the variable enters the model or not (for each value of $\lambda$). Then, the algorithm iterates over a regularization path of decreasing values of $\lambda$, which progressively enables variables to enter the model (Figure  \ref{fig:CMSA}). So, in the second step, the number of variables is increasing progressively, and at some point the computation stops due to the early stopping criterion (when prediction is getting worse on the corresponding validation set, see figure \ref{fig:CMSA}).

@@

Regarding memory requirements, this is basically the same as for computation time. Indeed, active variables are accessed in memory thanks to memory-mapping when they are used \cite[]{prive2017efficient}. 
If there are too many variables accessed, the operating system (OS) has to make some space for new coming variables. 
There should never be any memory issue thanks to memory-mapping. Yet, at some point, if there are too many active variables, the OS would regularly swap memory between disk and RAM, which is very slow. 
If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM. 
We have quickly tested processing each chromosome separately and then combining the models later but this did not perform as well as fitting all SNPs at once (data not shown). 
Another widely used approach is to apply the penalized regression on a subset of SNPs using univariate tests (GWAS) to prioritize SNPs. Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm here by achieving a correlation of 65.5\% for height using 350K individuals of the UK Biobank, whereas \cite{accuheight2018} achieved only 61\% for the same sample size when using only the 50K most significantly associated SNPs [TODO: COMPARER PLUTOT A MOI DIRECTEMENT].

@@

When there are many effects of small size and a small sample size, PLR performs worse than (the best result for) C+T. This may be because PLR is too conservative in including effects, whereas it has been shown in the literature that, in such underpowered scenarios, including (nearly) all SNPs was the best strategy \cite[]{purcell2009common}.
In scenario \textnumero3, by varying sample size, we show that when increasing sample size, PLR achieves higher predictive performance than C+T at some point and that the benefits of using PLR over C+T increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}). Moreover, in scenario \textnumero2, we decrease the search space (p: number of SNPs) in order to increase the polygenicity of the models and to virtually increase the sample size \cite[]{dudbridge2013power}. In this scenario, even when there are small effects and a high polygenicity (3000 causal variants out of 18,941), PLR gets a large increase in predictive performance, now consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
}
